Probabilistic Model of Bilateral Lymphatic Spread in Head and Neck Cancer

### 2024-09-04 _tex/manuscript.tex
Built site for gh-pages

\caption{}\label{T_b5301}\tabularnewline
\toprule\noalign{}
Institution & Total & Age (median) & Neck Dissection & N0 & Early T-Cat.
& Mid. Ext. \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
Institution & Total & Age (median) & Neck Dissection & N0 & Early T-Cat.
& Mid. Ext. \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
Centre Léon Bérard & 325 & 60 & 100\% & 19\% & 69\% & 18\% \\
Inselspital Bern & 74 & 61 & 100\% & 18\% & 66\% & 14\% \\
University Hospital Zurich & 287 & 66 & 26\% & 18\% & 52\% & 31\% \\
Vall d\textquotesingle Hebron Barcelona Hospital & 147 & 58 & 5\% & 21\%
& 34\% & 34\% \\

@@

The three risk factors for contralateral involvement, midline extension,
T-category, and ipsilateral involvement are correlated. For example,
midline extension was present in 48\% of advanced T-category tumors but
only in 7\% of early T-category tumors. The higher prevalence of
contralateral metastases in advanced T-category tumors may thus be
partially explained by a higher rate of midline extension in these
tumors. However, the bottom right panel of figure~\ref{fig-data-strat}
indicates that advanced T-category and severe ipsilateral involvement do
represent additional risk factors. The figure considers only patients
with lateralized tumors not extending over the midline. For early
T-category and healthy ipsilateral levels I-V, only 1\% (1 out of 84
patients) shows involvement of contralateral level II. This increases to
5\% (9 out of 165 patients) if ipsilateral level II is involved, to 15\%
(9 out of 60 patients) if ipsilateral levels II and III are involved,
and to 23\% (6 out of 26 patients) for advanced T-category with
ipsilateral levels II and III involved.

@@

For example, we may be interested in the prevalence of contralateral LNL
II involvement (i.e., contra LNL II ``involved'' and all other LNLs
``masked'') under the scenario of early T-category (T0-T2) and no
midline extension (\(\epsilon=\texttt{False}\)). To compute this
prevalence in the data, we select all patients of this scenario (in our
data, this amounts to 379 patients). Of those, 28 were found to harbor
metastases in their contralateral LNL II. Therefore, the prevalence is
7\%.

@@

\caption{}\label{T_6f0e6}\tabularnewline
\toprule\noalign{}
Parameter & Mean & Std. Dev. \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
Parameter & Mean & Std. Dev. \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
Mid. ext. probability & 8.16\% & ± 0.48\% \\
ipsi: T ➜ I & 2.80\% & ± 0.26\% \\
ipsi: T ➜ II & 34.89\% & ± 1.40\% \\
ipsi: T ➜ III & 5.45\% & ± 0.66\% \\
ipsi: T ➜ IV & 0.94\% & ± 0.18\% \\
ipsi: T ➜ V & 1.83\% & ± 0.22\% \\
ipsi: T ➜ VII & 2.32\% & ± 0.26\% \\
contra: T ➜ I & 0.29\% & ± 0.09\% \\
contra: T ➜ II & 2.46\% & ± 0.29\% \\
contra: T ➜ III & 0.14\% & ± 0.07\% \\
contra: T ➜ IV & 0.19\% & ± 0.08\% \\
contra: T ➜ V & 0.05\% & ± 0.04\% \\
contra: T ➜ VII & 0.50\% & ± 0.17\% \\
Mixing ⍺ & 33.87\% & ± 4.32\% \\
I ➜ II & 62.50\% & ± 16.69\% \\
II ➜ III & 14.23\% & ± 1.64\% \\
III ➜ IV & 15.86\% & ± 1.93\% \\
IV ➜ V & 14.58\% & ± 3.76\% \\
late T-cat. binom. prob. & 44.98\% & ± 1.99\% \\

@@

In this subsection we illustrate parts of the mathematical framework
defined in the previous section. In figure~\ref{fig-model-midext-evo}
(top panel), we visualize the prior distribution over diagnose times
\(P(t)\). By parameter choice, early T-category tumors are diagnosed on
average after 3 time steps. Advanced T-category tumors are diagnosed on
average after 4.5 time steps (\(p_\text{adv.}=0.45\)) , a parameter
learned from the data to describe the higher involvement for advanced
T-category tumors. The probability per time step for the tumor to grow
over the midline, \(p_\epsilon = 0.0816\) yields the probability of
midline extension \(P(\epsilon \mid t)\) at a given time step \(t\) (red
line). The parameter is adjusted such that the model yields the correct
proportion of patients with tumors extending over the midline seen in
the dataset. The bottom panel in figure~\ref{fig-model-midext-evo} shows
the joint probability \(P(\epsilon, t)\) to be diagnosed at time step
\(t\) with a given state of midline extension and T-category.

@@

Figure~\ref{fig-model-prevalences-overall} shows nicely that the model
is capable of accurately accounting for the most important risk factors,
i.e.~T-category and midline extension. As observed in the data, the
model predicts that the prevalence of contralateral LNL II involvement
increases from below 8\% for early T-category lateralized tumors to
almost 40\% when the tumor is of advanced T-category and crosses the
mid-sagittal line. Similarly, the prevalence of contralateral LNL III
involvement increases from around 2\% for early T-category lateralized
tumors to almost 15\% when the tumor is of advanced T-category and
crosses the mid-sagittal line.

@@

Figure~\ref{fig-model-risks} (left panel) shows the predicted risk of
occult disease in contralateral LNL II. The most important variable
impacting the prediction for contralateral level II involvement in our
model is the tumor's lateralization. A patient with a clearly
lateralized early T-category tumor and a clinically N0 neck is predicted
to have a 1-2\% risk for occult disease in contralateral LNL II. For a
patient with a clearly lateralized early T-category tumor but
\emph{with} mid-sagittal extension, the risk increases to almost 7\%.

Also advanced T-category increases the risk of occult disease but plays
a lesser role: Considering the scenario of a tumor that crosses the
midline and an ipsilateral neck where at least LNL III is clinically
involved, the risk for occult contralateral LNL II disease is around
8.5\%. For the same scenario and an advanced T-category tumor, the risk
increases to 11\%.

Lastly, the predicted risk is also correlated via the time-steps to the
degree of ipsilateral involvement. Changing the aforementioned scenario
(advanced T-category, midline extension, ipsilateral LNL III clinically
involved) to one where the patient presents with a clinically N0 neck,
the risk for occult disease in the contralateral LNL II falls from 11\%
to 9.5\%.

Taken together, T-category and ipsilateral involvement may still
considerably impact the risk prediction for contralateral involvement:
In figure~\ref{fig-model-risks} the scenarios underlying the orange
(6.5\% ) and the red (11\% ) histograms differ in T-category (early vs
advanced) and clinical diagnosis (N0 vs ipsilateral at least LNL III
involved).

@@

As shown in the center subplot of figure~\ref{fig-model-risks}, the risk
for occult disease in the contralateral LNL III may only cross a 5\%
threshold if the upstream level II is clinically involved, too. If the
clinical diagnosis of the contralateral upstream level II is
pathologically confirmed by FNA the risk increases by another percentage
point. When the tumor additionally crosses the mid-sagittal line, the
risk may reach around 7\%.

@@

Our model's prediction for the contralateral LNL IV show that even in
the most extreme case, with an advanced T-category tumor extending over
the mid-sagittal line as well as metastases in every ipsilateral LNL and
every contralateral upstream level, the risk stays just below 3\%. We
only observe a risk of 5-6\% when the involvement of the upstream level
III is pathologically confirmed after taking a biopsy using FNA. This is
because the model's prior probability for contralateral LNL IV
involvement is so low that even given a clinical diagnosis in that
level, a false positive in LNL III is a more likely explanation. This
possibility, however, drops virtually to zero when tumor cells are
pathologically confirmed in level III.

@@

\begin{itemize}
\tightlist
\item
  LNL II when the tumor extends over the mid-sagittal line
\item
  LNL III when LNL II is clinically involved, may need confirmation via
  FNA
\item
  LNL IV only when upstream level III pathologically confirmed
\end{itemize}

@@

\caption{}\label{T_a95ce}\tabularnewline
\toprule\noalign{}
T-cat. & ipsi & Mid. ext. & \multicolumn{2}{l}{%
I} & \multicolumn{2}{l}{%
II} & \multicolumn{2}{l}{%
III} & \multicolumn{2}{l}{%
IV} & total \\
& & & n & \% & n & \% & n & \% & n & \% & n \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
T-cat. & ipsi & Mid. ext. & \multicolumn{2}{l}{%
I} & \multicolumn{2}{l}{%
II} & \multicolumn{2}{l}{%
III} & \multicolumn{2}{l}{%
IV} & total \\
& & & n & \% & n & \% & n & \% & n & \% & n \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
early & 0 & False & 0 & 0.00 & 1 & 1.16 & 0 & 0.00 & 0 & 0.00 & 86 \\
early & 0 & True & 0 & 0.00 & 1 & 10.00 & 1 & 10.00 & 0 & 0.00 & 10 \\
early & 1 & False & 1 & 0.53 & 11 & 5.82 & 2 & 1.06 & 1 & 0.53 & 189 \\
early & 1 & True & 1 & 11.11 & 2 & 22.22 & 0 & 0.00 & 0 & 0.00 & 9 \\
early & ≥ 2 & False & 1 & 0.96 & 15 & 14.42 & 3 & 2.88 & 3 & 2.88 &
104 \\
early & ≥ 2 & True & 0 & 0.00 & 3 & 30.00 & 4 & 40.00 & 1 & 10.00 &
10 \\
advanced & 0 & False & 0 & 0.00 & 2 & 5.88 & 0 & 0.00 & 0 & 0.00 & 34 \\
advanced & 0 & True & 0 & 0.00 & 3 & 12.50 & 0 & 0.00 & 0 & 0.00 & 24 \\
advanced & 1 & False & 0 & 0.00 & 3 & 4.55 & 0 & 0.00 & 0 & 0.00 & 66 \\
advanced & 1 & True & 1 & 1.64 & 18 & 29.51 & 5 & 8.20 & 1 & 1.64 &
61 \\
advanced & ≥ 2 & False & 4 & 5.26 & 16 & 21.05 & 7 & 9.21 & 4 & 5.26 &
76 \\
advanced & ≥ 2 & True & 3 & 3.80 & 46 & 58.23 & 18 & 22.78 & 8 & 10.13 &
79 \\

>>>>

\caption{}\label{T_8749e}\tabularnewline
\toprule\noalign{}
Institution & Total & Age (median) & Neck Dissection & N0 & Early T-Cat.
& Mid. Ext. \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
Institution & Total & Age (median) & Neck Dissection & N0 & Early T-Cat.
& Mid. Ext. \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
Centre Léon Bérard & 325 & 60 & 100\% & 19\% & 69\% & 18\% \\
Inselspital Bern & 74 & 61 & 100\% & 18\% & 66\% & 14\% \\
University Hospital Zurich & 287 & 66 & 26\% & 18\% & 52\% & 31\% \\
Vall d\textquotesingle Hebron Barcelona Hospital & 147 & 58 & 5\% & 21\%
& 34\% & 34\% \\

@@

The three risk factors for contralateral involvement, midline extension,
T-category, and ipsilateral involvement are correlated. For example,
midline extension was present in 45.56\% of advanced T-category tumors
but only in 6.13\% of early T-category tumors. The higher prevalence of
contralateral metastases in advanced T-category tumors may thus be
partially explained by a higher rate of midline extension in these
tumors. However, the bottom right panel of figure~\ref{fig-data-strat}
indicates that advanced T-category and severe ipsilateral involvement do
represent additional risk factors. The figure considers only patients
with lateralized tumors not extending over the midline. For early
T-category and healthy ipsilateral levels I-V, only 1.85\% (2 out of 108
patients) shows involvement of contralateral level II. This increases to
6.73\% (14 out of 208 patients) if ipsilateral level II is involved, to
14.29\% (10 out of 70 patients) if ipsilateral levels II and III are
involved, and to 39.68\% (25 out of 63 patients) for advanced T-category
with ipsilateral levels II and III involved.

@@

For example, we may be interested in the prevalence of contralateral LNL
II involvement (i.e., contra LNL II ``involved'' and all other LNLs
``masked'') under the scenario of early T-category (T0-T2) and no
midline extension (\(\epsilon=\texttt{False}\)). To compute this
prevalence in the data, we select all patients of this scenario (in our
data, this amounts to 379 patients). Of those, 27 were found to harbor
metastases in their contralateral LNL II. Therefore, the prevalence is
7.12\%.

@@

\caption{}\label{T_0105a}\tabularnewline
\toprule\noalign{}
Parameter & Mean & Std. Dev. \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
Parameter & Mean & Std. Dev. \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
Mid. ext. probability & 8.16\% & ± 0.48\% \\
ipsi: T ➜ I & 2.80\% & ± 0.26\% \\
ipsi: T ➜ II & 34.89\% & ± 1.40\% \\
ipsi: T ➜ III & 5.45\% & ± 0.66\% \\
ipsi: T ➜ IV & 0.94\% & ± 0.18\% \\
ipsi: T ➜ V & 1.83\% & ± 0.22\% \\
ipsi: T ➜ VII & 2.32\% & ± 0.26\% \\
contra: T ➜ I & 0.29\% & ± 0.09\% \\
contra: T ➜ II & 2.46\% & ± 0.29\% \\
contra: T ➜ III & 0.14\% & ± 0.07\% \\
contra: T ➜ IV & 0.19\% & ± 0.08\% \\
contra: T ➜ V & 0.05\% & ± 0.04\% \\
contra: T ➜ VII & 0.50\% & ± 0.17\% \\
Mixing ⍺ & 33.87\% & ± 4.32\% \\
I ➜ II & 62.50\% & ± 16.69\% \\
II ➜ III & 14.23\% & ± 1.64\% \\
III ➜ IV & 15.86\% & ± 1.93\% \\
IV ➜ V & 14.58\% & ± 3.76\% \\
late T-cat. binom. prob. & 44.98\% & ± 1.99\% \\

@@

In this subsection we illustrate parts of the mathematical framework
defined in the previous section. In figure~\ref{fig-model-midext-evo}
(top panel), we visualize the prior distribution over diagnose times
\(P(t)\). By parameter choice, early T-category tumors are diagnosed on
average after 3 time steps. Advanced T-category tumors are diagnosed on
average after 4.5 time steps, because the parameter \(p_\text{adv.}\)
was learned to be 44.98\% for advanced T-category tumors. The
probability per time step for the tumor to grow over the midline
(\(p_\epsilon\)) was on average 8.16\%. Using this, one can compute the
probability of midline extension \(P(\epsilon \mid t)\) at a given time
step \(t\) (red line in figure~\ref{fig-model-midext-evo}). The bottom
panel in this figure shows the joint probability \(P(\epsilon, t)\) to
be diagnosed at time step \(t\) with a given state of midline extension
and T-category.

@@

Figure~\ref{fig-model-prevalences-overall} shows nicely that the model
is capable of accurately accounting for the most important risk factors,
i.e.~T-category and midline extension. As observed in the data, the
model predicts that the prevalence of contralateral LNL II involvement
increases from 7.1\% for early T-category lateralized tumors to 39.17\%
when the tumor is of advanced T-category and crosses the mid-sagittal
line. Similarly, the prevalence of contralateral LNL III involvement
increases from around 1.55\% for early T-category lateralized tumors to
almost 14.25\% when the tumor is of advanced T-category and crosses the
mid-sagittal line.

@@

Figure~\ref{fig-model-risks} (left panel) shows the predicted risk of
occult disease in contralateral LNL II. The most important variable
impacting the prediction for contralateral level II involvement in our
model is the tumor's lateralization. A patient with a clearly
lateralized early T-category tumor and a clinically N0 neck is predicted
to have a 1-2\% risk for occult disease in contralateral LNL II. For a
patient with a clearly lateralized early T-category tumor but
\emph{with} mid-sagittal extension, the risk increases to 7.58\%.

Also advanced T-category increases the risk of occult disease but plays
a lesser role: Considering the scenario of an early T-category tumor
that crosses the midline and an ipsilateral neck where the LNLs II and
III are clinically involved, the risk for occult contralateral LNL II
disease is around 9.06\%. For the same scenario and an advanced
T-category tumor, the risk increases to 11.25\%.

Lastly, the predicted risk is also correlated via the time-steps to the
degree of ipsilateral involvement. When an early T-category patient with
a tumor that extends over the mid-sagittal line presents with a
clinically N0 ipsilateral neck instead of involvement in the levels II
and III, the risk is 7.58\% instead of 9.06\%.

Taken together, T-category and ipsilateral involvement may still
considerably impact the risk prediction for contralateral involvement:
In figure~\ref{fig-model-risks} the scenarios underlying the orange
(7.58\%) and the purple (11.25\%) histograms differ in T-category (early
vs advanced) and clinical diagnosis (N0 vs ipsilateral LNL II and III
involved).

@@

As shown in the center subplot of figure~\ref{fig-model-risks}, the risk
for occult disease in the contralateral LNL III may only cross a 5\%
threshold if the upstream level II is clinically involved, too. If the
clinical diagnosis of the contralateral upstream level II is
pathologically confirmed by FNA the risk increases by another percentage
point to 5.58\%. When the tumor additionally crosses the mid-sagittal
line, the risk may reach around 6.94\%.

@@

Our model's prediction for the contralateral LNL IV show that even in
the most extreme case, with an advanced T-category tumor extending over
the mid-sagittal line as well as metastases in the ipsilateral LNLs
II-IV and the contralateral upstream levels II and III, the risk stays
at just 2.81\%. We only observe a risk of 5.68\% when the involvement of
the upstream level III is pathologically confirmed after taking a biopsy
using FNA. This is because the model's prior probability for
contralateral LNL IV involvement is so low that even given a clinical
diagnosis in that level, a false positive in LNL III is a more likely
explanation. This possibility, however, drops virtually to zero when
tumor cells are pathologically confirmed in level III.

\subsection{Contralateral LNLs I, V, and
VII}\label{contralateral-lnls-i-v-and-vii}

Even for advanced T-category patients whose tumor crosses the midline
and who present with extensive ipsi- and contralateral clinical
involvement, the risk for involvement in the LNLs I and VII never
exceeds 3\%. LNL V contralaterally also shows only a very low risk for
occult disease, but it does rise to 5.49\% in the same extreme case when
additionally the upstream level IV involvement has been confirmed with
FNA.

@@

\begin{itemize}
\tightlist
\item
  LNL II when the tumor extends over the mid-sagittal line
\item
  LNL III when LNL II is clinically involved, may need confirmation via
  FNA
\item
  LNL IV only when upstream level III pathologically confirmed
\item
  LNLs I and VII only when clinically involved
\item
  LNL V only in extreme cases (adv. T-cat., midext, all ipsi LNLs
  metastatic, contra II, III, and IV clinically involved plus cIV via
  FNA confirmed)
\end{itemize}

@@

\caption{}\label{T_a035c}\tabularnewline
\toprule\noalign{}
T-cat. & ipsi & Mid. ext. & \multicolumn{2}{l}{%
I} & \multicolumn{2}{l}{%
II} & \multicolumn{2}{l}{%
III} & \multicolumn{2}{l}{%
IV} & total \\
& & & n & \% & n & \% & n & \% & n & \% & n \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
T-cat. & ipsi & Mid. ext. & \multicolumn{2}{l}{%
I} & \multicolumn{2}{l}{%
II} & \multicolumn{2}{l}{%
III} & \multicolumn{2}{l}{%
IV} & total \\
& & & n & \% & n & \% & n & \% & n & \% & n \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
early & 0 & False & 0 & 1 & 0 & 0 & 0.00 & 1.16 & 0.00 & 0.00 & 86 \\
early & 0 & True & 0 & 1 & 1 & 0 & 0.00 & 10.00 & 10.00 & 0.00 & 10 \\
early & 0 & nan & 0 & 0 & 0 & 0 & 0.00 & 0.00 & 0.00 & 0.00 & 12 \\
early & 1 & False & 1 & 11 & 2 & 1 & 0.53 & 5.82 & 1.06 & 0.53 & 189 \\
early & 1 & True & 1 & 2 & 0 & 0 & 11.11 & 22.22 & 0.00 & 0.00 & 9 \\
early & 1 & nan & 0 & 3 & 0 & 1 & 0.00 & 9.68 & 0.00 & 3.23 & 31 \\
early & ≥ 2 & False & 1 & 15 & 3 & 3 & 0.96 & 14.42 & 2.88 & 2.88 &
104 \\
early & ≥ 2 & True & 0 & 3 & 4 & 1 & 0.00 & 30.00 & 40.00 & 10.00 &
10 \\
early & ≥ 2 & nan & 0 & 3 & 0 & 0 & 0.00 & 13.64 & 0.00 & 0.00 & 22 \\
advanced & 0 & False & 0 & 2 & 0 & 0 & 0.00 & 5.88 & 0.00 & 0.00 & 34 \\
advanced & 0 & True & 0 & 3 & 0 & 0 & 0.00 & 12.50 & 0.00 & 0.00 & 24 \\
advanced & 0 & nan & 0 & 0 & 0 & 0 & 0.00 & 0.00 & 0.00 & 0.00 & 3 \\
advanced & 1 & False & 0 & 3 & 0 & 0 & 0.00 & 4.55 & 0.00 & 0.00 & 66 \\
advanced & 1 & True & 1 & 18 & 5 & 1 & 1.64 & 29.51 & 8.20 & 1.64 &
61 \\
advanced & 1 & nan & 0 & 0 & 0 & 0 & 0.00 & 0.00 & 0.00 & 0.00 & 9 \\
advanced & ≥ 2 & False & 4 & 16 & 7 & 4 & 5.26 & 21.05 & 9.21 & 5.26 &
76 \\
advanced & ≥ 2 & True & 3 & 46 & 18 & 8 & 3.80 & 58.23 & 22.78 & 10.13 &
79 \\
advanced & ≥ 2 & nan & 0 & 0 & 0 & 0 & 0.00 & 0.00 & 0.00 & 0.00 & 8 \\
